Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-06-05 PhDThesis.tex
finished MCMC example. One more example to go before next section

\subsection{Poisson process}

@@

%TODO: JK69 as example
% \subsection{Example: JK69 substitution model}

@@

We need to note several important features of the algorithm.
The most important part of the algorithm is the proposal density. 
Given the current state it proposes the next state independent of the past states.
We can recall that this is the Markov property, in fact the output of the algorithm generates a Markov chain (hence the name).
Metropolis-Hastings algorithm allows for the use of asymetrical proposal densities such that $q\left(\theta\left[t\right]|\theta^{*}\right)\neq q\left(\theta^{*}|\theta\left[t\right]\right)$.
The acceptance probability is simply the ratio of the target posterior evaluated at the proposal and at the current state of the chain, times the ratio of the proposal: 
% prior ratio times the likelihood ratio times the proposal ratio:

@@

% TODO: better title
\subsection{Example}

%TODO: explain probabilities

%---METROPOLIS PATH---%
\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{metropolis} 
\end{center}
\caption{{ \footnotesize {\bf AAA.} 
}}
\label{fig:metropolis}
\end{figure}

%%%
\section{Bayesian phylogeography}

>>>>

\subsection{Poisson process\label{sub:poisson}}

@@

%TODO: JK69 as example
\subsection{Example: Jukes and Cantor substitution model\label{sub:jc69}}

\ref{Jukes1969}

TODO

@@

We need to note several important features of the algorithm.
The most important part of the algorithm is the proposal density. 
Given the current state it proposes the next state independent of the past states.
We can recall that this is the Markov property, in fact the output of the algorithm generates a Markov chain (hence the name).
Metropolis-Hastings algorithm allows for the use of asymetrical proposal densities such that $q\left(\theta\left[t\right]|\theta^{*}\right)\neq q\left(\theta^{*}|\theta\left[t\right]\right)$.
The acceptance probability is simply the ratio of the target posterior evaluated at the proposal and at the current state of the chain, times the ratio of the proposal: 
% prior ratio times the likelihood ratio times the proposal ratio

@@

% TODO: better title, show how to derive the probabilities (in JC69 chapter)
\subsection{Example: Estimating genetic distance}

Let us revisit the toy example in which we compare two sequences and find $x=10$ differences among $n=100$ sites.
This time however we will use MCMC to estimate the genetic distance between the two sequences using Jukes and Cantor model \ref{Jukes1969} of nucleotide substitution (see subsection \ref{sub:jc69}).
Genetic distance is defined as the expected number of substitutions of nucleotide base pairs per site, which to match the convention form previous sections we will denote as $\theta=rt$, where $r$ is the substitution rate (see subsection \ref{sub:poisson}) and $t$ is the waiting time between the two sequences.
The time and rate are typically confounded, therefore they are by convention estimated as a single parameter of interest $\theta$.
From subsection \ref{sub:jc69} we know that under the JC69 model the probability the probability that a single site is different between two sequences separated by distance $\theta$ is given by:

\begin{equation}
p=\frac{3}{4}\left(1-e^{\left(-4/3\right)\theta}\right).
\label{eq:distance1}
\end{equation}

\noindent
From this we can model the likelihood of observing $x$ differences out of $n$ sites using the binomial model:

\begin{equation}
P\left(\mathbf{X}|\theta\right)=L\left(\mathbf{X}|\theta\right)=\left(\begin{array}{c}
n\\
k
\end{array}\right)p^{x}(1-p)^{n-x},
\label{eq:likelihood1}
\end{equation}

\noindent
where $p$ is as defined by (\ref{eq:distance1}).
We use an exponential prior on $\theta$ with mean $1/\lambda$:

\begin{equation}
P\left(\theta,\lambda\right)=\frac{1}{\lambda}e^{-(1/\lambda)\theta},
\label{eq:expPrior}
\end{equation}

\noindent
and set $\lambda=10$.
For the proposal we choose a jumping kernel with a windows size of $w=0.1$.

%---METROPOLIS PATH---%
\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{metropolis} 
\end{center}
\caption{{ \footnotesize {\bf MCMC output for estimating sequence disctance under JC69 model of nucleotide evolution..} 
}}
\label{fig:metropolis}
\end{figure}

Figure \ref{fig:metropolis} shows $1000$ iterations of the chain generated by Algorithm \ref{alg:metropolisHastings} in the upper row.
The red line indicates the maximum likelihood estimate $\hat{\theta}$ of parameter $\theta$ which is calculated by setting $\frac{d\ell\left(\mathbf{X}|\theta\right)}{d\theta}=0$, where $l\left(\mathbf{X}|\theta\right)$ is the logarithm of likelihood defined by (\ref{eq:likelihood1}):

\begin{equation}
\hat{\theta}=\frac{3}{4}log\left(1-\frac{4}{3}\cdot\frac{x}{n}\right)
\label{eq:mle}
\end{equation}

In the second row of Figure \ref{fig:metropolis} we can see an estimate of the posterior density, along with the sample mean.
The acceptance rate is high, suggesting that almost as many steps are accepted as are rejected.

\section{Bayesian phylogeography}
